The brightness and spatial distributions of terrestrial radio sources

### 2013-01-18 articles/Article-RFI-distribution/article-rfi-distribution.tex
Dealt with comments from ref II

Converting from a subtracted term in decibels to a flux density factor $L_S$, and collecting the terms of Eq.~\eqref{eq:hata-model} that are not depending on $r$ in a single variable $\zeta$, results in
\begin{equation} \label{eq:propagation-loss-factor}
 L_S = \zeta r^{4.49 - 0.655 \log_{10} h_b},
\end{equation}
with
\begin{equation} \label{eq:zeta-definition}
 \zeta = \frac{f_c^{2.616}}{h_b^{1.382}} - 10^{6.955-\frac{1}{10} a(h_m)}.
\end{equation}
Note that according to Hata's model, the exponent of the power law in Eq.~\eqref{eq:propagation-loss-factor} depends only on the height of the transmitting antenna, i.e., it is independent of frequency, receiver height and urban density. Now, if in Eqs.~\eqref{eq:cumulative-rfi-distribution} and \eqref{eq:two-dimensional-distribution} one replaces the definition of $S(r)$ from Eq.~\eqref{eq:amplitude-fall-off-in-free-space} with one that includes the propagation effects,
\begin{equation}
 S(r) = \frac{\zeta Ig}{r^\eta},
\end{equation}
with $\eta = 4.49 - 0.655 \log_{10} h_b$, one finds the frequency density function $f_p$ that considers propagation effects,
% S(r):=I*g/(r^alpha);
% assume(g>0,I>0,S>0);
% diff(rho*%pi*r^2,S),solve(S(r)=S,r);
\begin{align} \label{eq:frequency-density-with-propagation}
f_p(S)
= & \frac{d}{dS} \left[ N - \rho \pi \left( \frac{\zeta Ig }{S} \right)^{2/\eta} \right]
= \frac{\rho 2\pi}{\eta S} \left( \frac{\zeta Ig}{S} \right)^{2/\eta}.
\end{align}
Consequently, due to non-free-space propagation effects, the observed log-log histogram is predicted to have a $-(\frac{2}{\eta}+1)$ slope. By substituting $\eta$, one finds
\begin{equation}
\textrm{slope}(h_b) = \frac{1}{0.3275 \log_{10} h_b - 2.245} - 1.
\end{equation}
\begin{figure}
\begin{center}\hspace{-1cm}\includegraphics[width=6cm]{img/transmitter-height/plot-transmitter-height-trimmed}
\caption{Effect of transmitter height on the slope of a log-log histogram.}
\label{fig:plot-transmitter-height}
\end{center}
\end{figure}
This is valid for transmitters that have a height of 30--200 m, the range over which Hata's model was defined. This yields estimated distribution slopes of $-1.57$ and $-1.67$ for $30$ m and $200$ m high transmitters respectively. In Figure~\ref{fig:plot-transmitter-height}, the slope value is plotted as a function of the transmitter height, including extrapolated values for transmitter heights down to 1 m.

@@

The real and imaginary components of receiver noise are independent and identically Gaussian distributed with zero mean and equal variance. Consequently, an amplitude $x$ will be Rayleigh distributed \citep[\S6-2]{papoulis-stochastic-processes}: 
\begin{equation}\label{eq:rayleigh-formula}
f_\textrm{noise}(x) =
\begin{cases}
\frac{x}{\sigma^2} e^{\frac{-x^2}{2\sigma^2}} & x > 0, \\
0 & \textrm{otherwise.}
\end{cases}
\end{equation}
Because most of the samples will be unaffected by RFI, this will be the dominating distribution. The Rayleigh distribution is plotted together with the -2 power-law distribution of Eq.~\eqref{eq:two-dimensional-distribution} in Fig.~\ref{fig:rayleigh-and-rfi-distributions}.

@@

If we assume that $I$, the average source strength; $g$, the average instrumental gain; and $r_L$, the distance over which those sources are visible, are all unknown, it can be seen from Eq.~\eqref{eq:drawing-combined-distribution} that given the histogram we can solve neither $r_L$, $I$ or $g$, as they all have the same effect of scaling the -2 power-law part of the histogram. Although calibration could in theory solve $g$, almost all sources will come in through the edges of the beam, and finding the expected values for the gains is therefore hard. The effects of these parameters on the histogram will be further discussed in \S\ref{sec:parameter-variability}.

@@

For completeness we show the formula for sampling apparent source strengths that include propagation effects according to Hata's model, which can be used for sampling realistic source amplitudes in simulations. This can be derived by integration, scaling and inversion of the frequency density function in Eq.~\eqref{eq:frequency-density-with-propagation}, resulting in:
\begin{equation} \label{eq:sample-with-propagation}
\tilde{S}_{\textrm{RFI}} \leftarrow \frac{\zeta I g}{x_u^{\eta/2} r_L^\eta} e^{i 2 \pi y_u}.
\end{equation}
Here, $\tilde{S}_{\textrm{RFI}}$ is a new complex RFI sample; $\zeta$ and $\eta$ are as defined below Eq.~\eqref{eq:zeta-definition}; $I$ is the average intrinsic strength; $g$ is the gain of the instrument; $0 \le x_u,y_u < 1$ are two independently drawn uniformly distributed samples; and $r_L$ is the maximum distance of visible sources.

\subsection{Parameter variability} \label{sec:parameter-variability}
In reality, the parameters $\rho$, $I$ and $g$, which are the source density per unit area, source strength and instrumental gain respectively, will not be constant over time and frequency but have a stochastic nature. However, since each specific value for these parameters produces a power law, the combined distribution will still show a power law, as long as the parameters follow a distribution that is steep at high amplitudes (in $\log$--$\log$ space), such as a Gaussian or uniform distribution.

@@

With estimates of $\alpha$, $\beta$, $S_L$ and $S_U$, one has obtained a parametrization of the RFI distribution. As was shown in \S\ref{sec:histogram-noise}, the left-most point where the power-law distribution falls off is $Ig/r^2$ -- assuming free space propagation for the moment -- and therefore $S_L = Ig/r^2$. For Hata's propagation model, the more generic solution $S_L = \zeta Ig / r^\alpha$ is found. This value represents the apparent brightness of the sources that are the furthest away from the telescope. With a fully parametrized distribution of the effect of RFI sources, an empirical model for RFI effects can be made. Moreover, one can calculate $E(S_R)$, the expected apparent strength of RFI:
\begin{align} \label{eq:expected-value-rfi-def}
E(S_R) & = \frac{1}{N_{LU}}\int\limits_{S_L}^{S_U} \beta S^\alpha S dS
 = \frac{\beta}{N_{LU}} \left[ \frac{1}{\alpha+2} S^{\alpha+2} \right]_{S_L}^{S_U} \\
 & = \frac{\beta \left( S_U^{\alpha+2} - S_L^{\alpha+2} \right) }{N_{LU} \left( \alpha+2\right) }
\end{align}
Here, $N_{LU}$ is the number of samples between $S_L$ and $S_U$ after normalizing for the bin size:
\begin{align} \label{eq:n-between-sl-su}
N_{LU} = \int\limits_{S_L}^{S_U} \beta S^\alpha dS.
\end{align}
Substitution and simplification of these two results in
\begin{align} \label{eq:expected-value-rfi}
 E(S_R) & = \frac
{\left( S_U^{\alpha+2} - S_L^{\alpha+2}\right)\left(\alpha+1\right)} 
{\left(S_U^{\alpha+1} - S_L^{\alpha+1}\right) \left(\alpha+2 \right)}.
\end{align}
This is essentially the flux density that is caused by RFI without using RFI detection or excision algorithms. $E(S_R)$ has the same units as $S_L$ and $S_U$, thus after calibration (see \S\ref{sec:calibration}) could be given in Jy. In practice, the increase of system noise after correlation is much less severe because of RFI flagging, which excises a part of the RFI. One can assume that all RFI above a certain power level is found by the detector. Since modern RFI detection algorithms can find all RFI that is detectable ``by eye'' \citep{post-correlation-rfi-classification}, this power level will be near the level of the noise mode. In
\citet{lofar-radio-environment}
the false-positives rate for the AOFlagger is estimated to be 0.5\%. An estimate of $S_d$, the average lower limit of detected RFI, can be calculated by finding the point on the distribution where the area under the distribution to the right of $S_d$ equals the real number (true positives) of RFI samples. Therefore, the limit is calculated similar to Eq.~\eqref{eq:lower-limit-1}, where the term $N_\textrm{part} - N_\textrm{total}$ needs to be replaced with $N_\textrm{RFI}$, which equals the total number of samples detected as RFI minus the 0.5\% false positives.

@@

\section{Conclusions and discussion} \label{sec:dist-discussion}
We have analysed the histogram of visibility amplitudes of LOFAR observations and found that, within a significant range of the histogram, the contribution of RFI sources follows a power-law distribution. The found power-law exponents of $-1.62$ and $-1.53$ for the 30--78~MHz LBA and 115--163~MHz HBA observations respectively, can be explained by a uniform spatial distribution of RFI sources, affected by propagation described surprisingly well by Hata's electromagnetic propagation model. Taken at face value these exponents imply in Hata's model that the average transmitting heights for sources affecting the LBA and HBA are $79$ and $13$~m respectively. Hata's model only goes down to 150 MHz, and it is possible that the electromagnetic fall-off due to propagation will be different for lower frequencies, e.g. because the effect of the ionosphere becomes stronger. Intervals for the exponents with representative $3\sigma$ confidence boundaries are $[-1.80;-1.44]$ for the LBA and $[-1.53;-1.49]$ for the HBA. The estimate of the HBA is thus more accurate, because its histogram deviates less from the power law.

@@

Finally, future RFI excision strategies will further enhance detection accuracy. Currently, RFI excision is applied only on the raw data from single interferometers at high resolution. Once data from a large number of nights are collected, it will be possible to detect and excise RFI more accurately, by looking at the averaged data from multiple nights and/or multiple interferometers. We have shown that the current detection algorithms can detect RFI well below the noise. Therefore, if data from different nights or interferometers are summed, and there is stationary RFI in the set, it will become detectable. All RFI that is below the noise but is not stationary, will act like normal noise and will therefore be harmless.

>>>>

Converting from a subtracted term in decibels to a flux density factor $L_S$, and collecting the terms of Eq.~\eqref{eq:hata-model} that are not depending on $r$ in a single variable $\zeta$, results in
\begin{equation} \label{eq:propagation-loss-factor}
 L_S = \zeta r^{4.49 - 0.655 \log_{10} h_b},
\end{equation}
with
\begin{equation} \label{eq:zeta-definition}
 \zeta = \frac{f_c^{2.616}}{h_b^{1.382}} - 10^{6.955-\frac{1}{10} a(h_m)}.
\end{equation}
Note that according to Hata's model, the exponent of the power law in Eq.~\eqref{eq:propagation-loss-factor} depends only on the height of the transmitting antenna, i.e., it is independent of frequency, receiver height and urban density. Now, if in Eqs.~\eqref{eq:cumulative-rfi-distribution} and \eqref{eq:two-dimensional-distribution} one replaces the definition of $S(r)$ from Eq.~\eqref{eq:amplitude-fall-off-in-free-space} with one that includes the propagation effects,
\begin{equation}
 S(r) = \frac{Ig}{\zeta r^\eta},
\end{equation}
with $\eta = 4.49 - 0.655 \log_{10} h_b$, one finds the frequency density function $f_p$ that considers propagation effects,
% S(r):=I*g/(r^alpha);
% assume(g>0,I>0,S>0);
% diff(rho*%pi*r^2,S),solve(S(r)=S,r);
\begin{align} \label{eq:frequency-density-with-propagation}
f_p(S)
= & \frac{d}{dS} \left[ N - \rho \pi \left( \frac{Ig }{\zeta S} \right)^{2/\eta} \right]
= \frac{\rho 2\pi}{\eta S} \left( \frac{Ig}{\zeta S} \right)^{2/\eta}.
\end{align}
Consequently, due to non-free-space propagation effects, the observed log-log histogram is predicted to have a $-(\frac{2}{\eta}+1)$ slope. By substituting $\eta$, one finds
\begin{equation}
\textrm{slope}(h_b) = \frac{1}{0.3275 \log_{10} h_b - 2.245} - 1.
\end{equation}
\begin{figure}
\begin{center}\hspace{-1cm}\includegraphics[width=6cm]{img/transmitter-height/plot-transmitter-height-trimmed}
\caption{Effect of transmitter height on the slope of a log-log histogram.}
\label{fig:plot-transmitter-height}
\end{center}
\end{figure}
This is valid for transmitters that have a height of 30--200 m, the range over which Hata's model was defined. This yields estimated distribution slopes of $-1.57$ and $-1.67$ for $30$ m and $200$ m high transmitters respectively. In Figure~\ref{fig:plot-transmitter-height}, the slope value is plotted as a function of the transmitter height, including extrapolated values for transmitter heights down to 1 m.

@@

The real and imaginary components of the noise in the cross-correlations are independent and identically Gaussian distributed with zero mean and equal variance. Consequently, an amplitude $x$ will be Rayleigh distributed \citep[\S6-2]{papoulis-stochastic-processes}: 
\begin{equation}\label{eq:rayleigh-formula}
f_\textrm{noise}(x) =
\begin{cases}
\frac{x}{\sigma^2} e^{\frac{-x^2}{2\sigma^2}} & x > 0, \\
0 & \textrm{otherwise.}
\end{cases}
\end{equation}
Because most of the samples will be unaffected by RFI, this will be the dominating distribution. The Rayleigh distribution is plotted together with the -2 power-law distribution of Eq.~\eqref{eq:two-dimensional-distribution} in Fig.~\ref{fig:rayleigh-and-rfi-distributions}.

@@

If we assume that $I$, the average source strength; $g$, the average instrumental gain; and $r_L$, the distance over which those sources are visible, are all unknown, it can be seen from Eq.~\eqref{eq:drawing-combined-distribution} that given the histogram we can solve neither $r_L$, $I$ or $g$, as they all have the same effect of scaling the -2 power-law part of the histogram. Although calibration could in theory solve $g$, almost all sources will come in through the far-sidelobes of the phased-array beam, and finding the expected values for the gains is therefore hard. The effects of these parameters on the histogram will be further discussed in \S\ref{sec:parameter-variability}.

@@

For completeness we show the formula for sampling apparent source strengths that include propagation effects according to Hata's model, which can be used for sampling realistic source amplitudes in simulations. This can be derived by integration, scaling and inversion of the frequency density function in Eq.~\eqref{eq:frequency-density-with-propagation}, resulting in:
\begin{equation} \label{eq:sample-with-propagation}
\tilde{S}_{\textrm{RFI}} \leftarrow \frac{I g}{\zeta x_u^{\eta/2} r_L^\eta} e^{i 2 \pi y_u}.
\end{equation}
Here, $\tilde{S}_{\textrm{RFI}}$ is a new complex RFI sample; $\zeta$ and $\eta$ are as defined below Eq.~\eqref{eq:zeta-definition}; $I$ is the average intrinsic strength; $g$ is the gain of the instrument; $0 \le x_u,y_u < 1$ are two independently drawn uniformly distributed samples; and $r_L$ is the maximum distance of visible sources.

\subsection{Parameter variability} \label{sec:parameter-variability}
In reality, the parameters $\rho$, $I$ and $g$, which are the source density per unit area, source strength and instrumental gain respectively, will not be constant, but can change over time and frequency. Therefore, they act like stochastic variables. However, since each specific value for these parameters produces a power law, the combined distribution will still show a power law, as long as the parameters follow a distribution that is steep at high amplitudes (in $\log$--$\log$ space), such as a Gaussian or uniform distribution.

@@

With estimates of $\alpha$, $\beta$, $S_L$ and $S_U$, one has obtained a parametrization of the RFI distribution. As was shown in \S\ref{sec:histogram-noise}, the left-most point where the power-law distribution falls off is $Ig/r^2$ -- assuming free space propagation for the moment -- and therefore $S_L = Ig/r^2$. For Hata's propagation model, the more generic solution $S_L = Ig / (\zeta r^\alpha)$ is found. This value represents the apparent brightness of the sources that are the furthest away from the telescope. With a fully parametrized distribution of the effect of RFI sources, an empirical model for RFI effects can be made. Moreover, one can calculate $E(S_R)$, the expected apparent strength of RFI:
\begin{align} \label{eq:expected-value-rfi-def}
E(S_R) & = \frac{1}{N_{LU}}\int\limits_{S_L}^{S_U} \beta S^\alpha S dS
 = \frac{\beta}{N_{LU}} \left[ \frac{1}{\alpha+2} S^{\alpha+2} \right]_{S_L}^{S_U} \\
 & = \frac{\beta \left( S_U^{\alpha+2} - S_L^{\alpha+2} \right) }{N_{LU} \left( \alpha+2\right) }
\end{align}
Here, $N_{LU}$ is the number of samples between $S_L$ and $S_U$ after normalizing for the bin size:
\begin{align} \label{eq:n-between-sl-su}
N_{LU} = \int\limits_{S_L}^{S_U} \beta S^\alpha dS.
\end{align}
Substitution and simplification of these two results in
\begin{align} \label{eq:expected-value-rfi}
 E(S_R) & = \frac
{\left( S_U^{\alpha+2} - S_L^{\alpha+2}\right)\left(\alpha+1\right)} 
{\left(S_U^{\alpha+1} - S_L^{\alpha+1}\right) \left(\alpha+2 \right)}.
\end{align}
This is essentially the flux density that is caused by RFI without using RFI detection or excision algorithms. $E(S_R)$ has the same units as $S_L$ and $S_U$, thus after calibration (see \S\ref{sec:calibration}) could be given in Jy. In practice, the increase of system noise after correlation is much less severe because of RFI flagging, which excises a part of the RFI. One can assume that all RFI above a certain power level is found by the detector. Since modern RFI detection algorithms can find all RFI that is detectable ``by eye'' \citep{post-correlation-rfi-classification}, this power level will be near the level of the noise mode. In
\citet{lofar-radio-environment}
the false-positives rate for the AOFlagger is estimated to be 0.5\%. An estimate of $S_d$, the average lower limit of detected RFI, can be calculated by finding the point on the distribution where the area under the distribution to the right of $S_d$ equals the real number (true positives) of RFI samples. Therefore, the limit is calculated similar to Eq.~\eqref{eq:lower-limit-1}, where the term $N_\textrm{part} - N_\textrm{total}$ needs to be replaced with $N_\textrm{RFI}$, which equals the total number of samples detected as RFI minus the 0.5\% false positives.

@@

\section{Conclusions and discussion} \label{sec:dist-discussion}
We have analysed the histogram of visibility amplitudes of LOFAR observations and found that, within a significant range of the histogram, the contribution of RFI sources follows a power-law distribution. The found power-law exponents of $-1.62$ and $-1.53$ for the 30--78~MHz LBA and 115--163~MHz HBA observations respectively, can be explained by a uniform spatial distribution of RFI sources, affected by propagation described surprisingly well by Hata's electromagnetic propagation model. Taken at face value these exponents imply in Hata's model that the average transmitting heights for sources affecting the LBA and HBA are $79$ and $13$~m respectively. There are no 79 m high transmitters nearby LOFAR stations in the LBA frequency range. Hata's model only goes down to 150 MHz, and it is possible that the electromagnetic fall-off due to propagation will be different for lower frequencies, e.g. because the effect of the ionosphere becomes stronger. Intervals for the exponents with representative $3\sigma$ confidence boundaries are $[-1.80;-1.44]$ for the LBA and $[-1.53;-1.49]$ for the HBA. The estimate of the HBA is thus more accurate, because its histogram deviates less from the power law.

@@

Finally, future RFI excision strategies will further enhance detection accuracy. Currently, RFI excision is applied only on the raw data from single interferometers at high resolution. Once data from a large number of nights are collected, it will be possible to detect and excise RFI more accurately, by looking at the averaged data from multiple nights and/or multiple interferometers. We have shown that the current detection algorithms can detect RFI well below the noise. Therefore, if data from different nights or interferometers are summed, and there is stationary RFI in the set, it will become detectable. All RFI that is at that point still below the noise, will act like normal noise and will therefore be harmless.
